Quantum H theorem, variational entropy, and relaxation processes

### 2021-01-25 tex/quantum.tex
Se agregó un breve desarrolo para la sección de relajación en términos de la funcional H evaluada en diferentes puntos. Corregí un texto en rojo en la sección quantum donde estaban mal identificados los signos para bosones y fermiones. También agregué en rojo la subsección OUT OF EQUILIBRIUM, NON-HOMOGENEOUS SYSTEMS debido a que ahí no se hace mención sobre sistemas fuera de equilibrio sino hasta que demostramos el teorema H.
Cambié lo de la subsección en rojo, error de numeración
Rewording: All sections finished but 5.

The classical $H$-theorem is even to these days one of the pillars whereon the classical statistical
physics is founded. Unfortunately, regardless of several attempts have been made \cite{bib:silva2010,bib:deroeck2006,bib:grabert1974,bib:han2015,bib:das2018,bib:vonneumann2010}, 
the generality of the classical $H$-theorem has no an equally robust quantum match. In this section, we shall
propose and analize an alternative quantum $H$-functional using the variational method.
We start by briefly outline a typical textbook demonstration of the quantum $H$-theorem
\cite{bib:tolman}, then, we present the analysis of our proposed $H$-functional.

@@

Consider a diluted gas of $N$ non-interacting quantum particles (either bosons or fermions),
contained by a vessel of volume
$V$, at some temperature $T$, and total energy E. Starting from the Boltzmann definition
of entropy, the quantum $H$ functional is
%
\begin{equation}
   H_T=-\ln G,
\end{equation}
%
where $G$ describes the total number of accessible quantum states of the gas that satisfy the
above mentioned conditions \cite{bib:tolman}.
The quantum $H$-theorem can be demostrated as follows.  $G$ can be divided into groups of neighboring states,
$g_k$, and a certain occupation numbers, $n_k$, can be associated with each of these
groups. Thus, the above functional takes the form:
%
\begin{equation}\label{eq:quantumh}
    H_T=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, 
\end{equation}
%
where the upper and lower signs are for bosons and fermions, respectively.
Thus the time derivative of Eq.~(\ref{eq:quantumh}) is
%
\begin{equation}\label{eq:quantumdHdt}
\frac{dH_T}{dt}=\sum_{\kappa}\left[\ln n_{\kappa}-\ln\left(g_{\kappa}\pm n_{\kappa}\right)\right]
\frac{dn_{\kappa}}{dt}.
\end{equation}
%

Assuming that the energy exchange between particles is produced by collisions between particles.
By using perturbation theory, the rate of change in the number of particles in a group $\kappa$ is
%
\begin{eqnarray}\label{eq:changen}
    \frac{d n_{\kappa}}{dt}&=&-\sum_{\lambda,(\mu \nu)}A_{\kappa\lambda,\mu\nu} n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})\nonumber \\
    &&+\sum_{\lambda,(\mu \nu)}A_{\mu\nu,\kappa\lambda} n_{\mu}n_{\nu}(g_{\kappa}\pm n_{\kappa})(g_{\lambda}\pm n_{\lambda}).
\end{eqnarray}
%
Here $A_{\kappa\lambda,\mu\nu}n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})$ is the
expected number of collisions per unit time, in which two particles will be moved from groups
($\kappa$, $\lambda$) to ($\mu$, $\nu$), and the tensor $A_{\kappa\lambda,\mu\nu}$ is given by
%
\begin{equation}\label{eq:amnkldef}
  A_{\kappa\lambda,\mu\nu}=\frac{4\pi^{2}}{h}\frac{|I_1\pm I_2|^2}{\Delta \epsilon}.
\end{equation}
%
In Eq.~(\ref{eq:amnkldef}), $\Delta\epsilon$ is the net energy change occuring during the
collision and $|I_1-I_2|^2=|V_{mn,kl}|^2$ where
$V_{mn,kl}$ is the element of the transition matrix of a binary collision.
It is important to remark that in deriving Eq.~(\ref{eq:changen}), the equal a priori
probabilities and the random a priori phases hypotheses were taken valid. The 
the random a priori phases hypothesis can be
considered an analogous to the molecular chaos hypothesis
\cite{bib:das2018} as the mechanism to introduce stochasticity into the system.

Substituting Eq.~(\ref{eq:changen}) into Eq.~(\ref{eq:quantumdHdt}) after some algebra, it can be proven that
%
\begin{equation}
    \frac{dH_T}{dt}\leq 0. \label{eq:H-theorem-tolman}
\end{equation}
%

@@

We propose the following functional as an alternative $H$-functional
for quantum non-homogeneous diluted gases:
%
\begin{eqnarray}\label{eq:qHdef}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\bigg.\nonumber \\
    &&\qquad\qquad\bigg.\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M.
\end{eqnarray}
%
Here, the upper and lower sign refer to \commr{bosons and fermions respectively}.
In addition, when needed, each cell will have an associated local chemical potential,
$\alpha_M$, and a local $H$-functional, which is defined by:
%
\begin{equation}\label{eq:qHMdef}
   \mathcal{H}_M(t)=\sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)
   \pm \Big(1 \mp f_{Mn}(t)\Big) \ln \Big(1 \mp f_{Mn}(t)\Big) \bigg] \de V_M.
\end{equation}
%
Therefore, $\mathcal{N}_M$ and $\mathcal{E}_M$ as functions of time are given by:
%
\begin{subequations}
\begin{equation}
    {\mathcal{N}}_M(t) = \sum_{n}f_{Mn}(t) \delta V_M
\end{equation}
%
and
%
\begin{equation}
\mathcal{E}_M(t) = \sum_{n}f_{Mn}(t)\epsilon_{n} \de V_M,
\end{equation}
\end{subequations}
%
which are, for the whole system, constrained by the microcanonical restrictions
%
\begin{subequations}\label{eq:qGlobRest}
\begin{equation}\label{eq:qGlobRestN}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\right]\delta  V_M
    =\sum_{M=1}^{K}\mathcal{N}_{M}(t)\delta V_M=N,
\end{equation}
%
and
%
\begin{equation}\label{eq:qGlobRestE}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\epsilon_{n}\right]\delta V_M
    =\sum_{M=1}^{K}\mathcal{E}_M(t)\delta V_M=E. 
\end{equation}
\end{subequations}
%

Applying the variational method to $\mathcal{H}$, and using the
Lagrange multipliers
$\{\alpha_M\}$ and $\{\beta_M\}$, it is straightforward to obtain
(see also the discussion of Eq.~(\ref{eq:deltaHpdeltafpj})):
%
\begin{equation}\label{eq:relation}
\ln \left(\frac{1\mp f_{Mn}(t)}{f_{Mn}(t)} \right)=-\alpha_M(t)-\beta_M(t) \epsilon_{n},
\end{equation}
%
and solving for $f_{Mn}(t)$ yields
%
\begin{equation}\label{eq:qfMn}
f_{Mn}(t)=\frac{1}{\exp\big(-\alpha_M(t)-\beta_M(t) \epsilon_{n}\big)\pm 1}.
\end{equation}
%
Thus, in this zero-order approximation the form of equilibrium distribution functions are conserved.

%============================================================
\subsubsection{Properties of $\mathcal{H}$ for systems in equilibrium}
%============================================================
If the system is in equilibrium (we will show that it will also be valid at $t\to\infty$), the temperature becomes homogeneous throughout
the complete system. Also, the local number of particles, the local
energy, and the Lagrange multipliers do not depend on the cell number and they should be homogeneous,
this is 
%
\begin{subequations}\label{eq:qEqRestrictions}
\begin{equation}
   {\mathcal{N}}_M(t\to\infty)\equiv \bar{\mathcal{N}}=N/K,
\end{equation}
%
%
\begin{equation}
	{\mathcal{E}}_M(t\to\infty)\equiv \bar{\mathcal{E}}=E/K,
\end{equation}
\end{subequations}
%
%
\begin{subequations}\label{eq:qEqAlphaBeta}
\begin{equation}
	\alpha_M=\bar\alpha,\quad\forall M,
\end{equation}
%
and
\begin{equation}
	\beta_M=\bar\beta, \quad\forall M.
\end{equation}
\end{subequations}
%
Substituting Eqs.~(\ref{eq:qEqAlphaBeta}) into Eq.~(\ref{eq:qfMn}) yields the distribution
function of each cell in equilibrium to be
%
\begin{equation}\label{eq:qfneq}
    \bar f_{Mn}=\bar f_n =\frac{1}{\exp\big(-\bar\alpha-\bar\beta \epsilon_n\big)\pm 1},\quad\forall M.
\end{equation}
%
By means of the above equation, we can recover the distribution function and the entropy of a diluted quantum gas in equilibrium as follows. Setting $\bar\alpha=\mu/kT$ and $\bar\beta=-1/kT$,
and substituting them in Eq.~(\ref{eq:qfneq}), it renders the Fermi-Dirac and Bose-Einstein
distribution functions:
%
\begin{equation}\label{eq:qfneqtwo}
    \bar{f}_{n}=\frac{1}{\exp\big(\frac{{\epsilon_n}-\mu}{kT}\big)\pm 1},
\end{equation}
%
and substituting Eq.~(\ref{eq:qfneq}) into the negative of Eq.(\ref{eq:qHdef}), 
the entropy of a quantum ideal gas is obtained
%
\begin{eqnarray}\label{eq:varEntropyDef}
      S&=&\sum_{M=1}^{K}
        \sum_n  \left[
        		\left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
           \commr{\ln \left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)}
         \right]\nonumber \\
      &&\quad\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}
         \left(1 \mp \frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
         \right] \delta V_M.
\end{eqnarray}
%
This quantity is what we refer to as ``variational entropy'',
and this name reflects the fact that it was obtained \textit{via}
the variational method. {\color{blue} We will use it at the end of this article to analyze relaxation processes.}

@@

For quantum systems, we also accept the validity of the local equilibrium hypothesis
for every cell in the system. This allows us to define non-homogeneous systems,
wherein thermodynamic quantities are well-defined on a cell-per-cell basis,
and in terms of the equilibrium properties, we have

@@

In the present context, $|\Delta_M|$ and $|\delta_M|$ are sufficiently large so they are not
fluctuations of the system, but enough small so that the local equilibrium hypothesis is
valid for $t>0$ (we set $t_0=0$, and $t_0$ is the initial time at which the system is prepared).
Therefore, it is reasonable to re-write the distribution functions as:
%
\begin{equation}\label{eq:qFirstOrd}
   f_{Mn}(t)=\bar{f}_n(1+g_{Mn}(t)),\quad
   1\gg|g_{Mn}(t)|,
\end{equation}
%
from which it follows, by substituting Eq.~(\ref{eq:qFirstOrd}) into
Eq.~(\ref{eq:restrictionoutside}), that $\Delta_M$ and $\delta_M$ satisfy:
%
\begin{subequations}
\begin{equation}
    \Delta_M(t)=\sum_n \bar{f}_n g_{nM}
\end{equation}
%
and
%
\begin{equation}
	\delta_M(t)=\sum_n  \bar{f}_n g_{nM}\epsilon_n.
\end{equation}
\end{subequations}
%

@@

%\begin{proof}
To proof the quantum $H$-theorem, start by taking the time-derivative of Eq. (\ref{eq:qHdef}):
%
\begin{equation}\label{eq:deltaH}
   \frac{d \mathcal{H} (t)}{dt}= \sum_n \sum_{M=1}^{K} \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M.
\end{equation}
%
Subsequently, substitute Eq.~(\ref{eq:qFirstOrd}) in the above equation to obtain:
%
\begin{eqnarray}\label{eq:cambioH1}
    \frac{d\mathcal{H} (t)}{dt}&=&
      \sum_n \sum_{M=1}^{K} \bar{f}_{n}\ln \left[
        \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})}
      \right]\dot{g}_{nM} \delta V_M \nonumber \\
    &=&\sum_n \sum_{M=1}^{K} \bar{f}_n \left \{
      \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}
      -\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}
    \right \}\delta V_M.
\end{eqnarray}
%
Approximate the logarithmic terms, corresponding to
Fermi and Bose gases, through a Taylor series around $\bar f_n g_{nM}=0$:
%
\begin{subequations}\label{eq:qlnApprox}
\begin{equation}\label{eq:qlnApproxFermions}
	\ln[1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]
    	\approx \ln[1\mp\bar{f}_n]\mp\frac{\bar{f}_n}{1\mp\bar{f}_{n}} g_{nM}
\end{equation}
%
and
%
\begin{equation}\label{eq:qlnApproxBosons}
    \ln [\bar{f}_n+\bar{f}_n g_{nM}] \approx \ln [\bar{f}_n]+ g_{nM},
\end{equation}
\end{subequations}
%
respectively. Eq.~(\ref{eq:qlnApproxFermions}) is valid because, for non-extremely degenerated Fermi gases, 
$1-\bar{f}_n \gg \bar{f}_n|g_{nM}|$, and Eq.~(\ref{eq:qlnApproxBosons}) is fulfilled because,
for Boson gases,
$1+\bar{f}_n \gg \bar{f}_n |g_{nM}|$ when $\bar{f}_n \gg \bar{f}_n |g_{nM}|$.

>>>>

The classical $H$-theorem is even to these days one of the pillars on which the classical statistical
physics is founded. Unfortunately, regardless of several attempts that have been made \cite{bib:silva2010,bib:deroeck2006,bib:grabert1974,bib:han2015,bib:das2018,bib:vonneumann2010}, 
the generality of the classical $H$-theorem has no equally robust quantum match. In this section, we shall
propose and analize an alternative quantum $H$-functional using the variational method.
We start by briefly outlining a typical textbook demonstration of the quantum $H$-theorem
\cite{bib:tolman}, and subsequently we present the analysis of our proposed $H$-functional.

@@

Consider a diluted gas of $N$ non-interacting quantum particles (either bosons or fermions),
contained by a vessel of volume
$V$, at some temperature $T$, and total energy E. Starting from the Boltzmann definition
of entropy, the quantum $H$ functional is
%
\begin{equation}
   H_T=-\ln G,
\end{equation}
%
where $G$ describes the total number of accessible quantum states of the gas that satisfy the
above conditions \cite{bib:tolman}.
The quantum $H$-theorem can be demostrated as follows.  $G$ can be divided into groups of neighboring states,
$g_k$, and certain occupation numbers, $n_k$, can be associated with each of these
groups. Thus, the above functional takes the form:
%
\begin{equation}\label{eq:quantumh}
    H_T=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, 
\end{equation}
%
where the upper and lower signs are for bosons and fermions, respectively.
Thus the time derivative of Eq.~(\ref{eq:quantumh}) is
%
\begin{equation}\label{eq:quantumdHdt}
\frac{dH_T}{dt}=\sum_{\kappa}\left[\ln n_{\kappa}-\ln\left(g_{\kappa}\pm n_{\kappa}\right)\right]
\frac{dn_{\kappa}}{dt}.
\end{equation}
%

Assuming that the energy exchange between particles is produced by collisions between particles,
and using perturbation theory, the rate of change in the number of particles in a group $\kappa$ is
%
\begin{eqnarray}\label{eq:changen}
    \frac{d n_{\kappa}}{dt}&=&-\sum_{\lambda,(\mu \nu)}A_{\kappa\lambda,\mu\nu} n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})\nonumber \\
    &&+\sum_{\lambda,(\mu \nu)}A_{\mu\nu,\kappa\lambda} n_{\mu}n_{\nu}(g_{\kappa}\pm n_{\kappa})(g_{\lambda}\pm n_{\lambda}).
\end{eqnarray}
%
Here $A_{\kappa\lambda,\mu\nu}n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})$ is the
expected number of collisions per unit time, in which two particles will be moved from groups
($\kappa$, $\lambda$) to ($\mu$, $\nu$), and the tensor $A_{\kappa\lambda,\mu\nu}$ is given by
%
\begin{equation}\label{eq:amnkldef}
  A_{\kappa\lambda,\mu\nu}=\frac{4\pi^{2}}{h}\frac{|I_1\pm I_2|^2}{\Delta \epsilon}.
\end{equation}
%
In Eq.~(\ref{eq:amnkldef}), $\Delta\epsilon$ is the net energy change occuring during the
collision and $|I_1-I_2|^2=|V_{mn,kl}|^2$, where
$V_{mn,kl}$ is the element of the transition matrix of a binary collision.
It is important to remark that in deriving Eq.~(\ref{eq:changen}), the equal a priori
probabilities and the random a priori phases hypotheses were taken as valid. The 
random a priori phases hypothesis can be
considered an analogue of the molecular chaos hypothesis
\cite{bib:das2018}, as it is the mechanism by which stochasticity is introduced into the system.

Substituting Eq.~(\ref{eq:changen}) into Eq.~(\ref{eq:quantumdHdt}), it is straightforward to prove that
%
\begin{equation}
    \frac{dH_T}{dt}\leq 0. \label{eq:H-theorem-tolman}
\end{equation}
%

@@

We propose the following functional as an alternative $H$-functional
for quantum non-homogeneous diluted gases:
%
\begin{eqnarray}\label{eq:qHdef}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\bigg.\nonumber \\
    &&\qquad\qquad\bigg.\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M.
\end{eqnarray}
%
Here, the upper and lower sign refer to fermions and bosons respectively.
In addition, when needed, each cell will have an associated local chemical potential,
$\alpha_M$, and a local $H$-functional, which is defined by:
%
\begin{equation}\label{eq:qHMdef}
   \mathcal{H}_M(t)=\sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)
   \pm \Big(1 \mp f_{Mn}(t)\Big) \ln \Big(1 \mp f_{Mn}(t)\Big) \bigg] \de V_M.
\end{equation}
%
Therefore, $\mathcal{N}_M$ and $\mathcal{E}_M$ as functions of time are given by:
%
\begin{subequations}
\begin{equation}
    {\mathcal{N}}_M(t) = \sum_{n}f_{Mn}(t) \delta V_M
\end{equation}
%
and
%
\begin{equation}
\mathcal{E}_M(t) = \sum_{n}f_{Mn}(t)\epsilon_{n} \de V_M,
\end{equation}
\end{subequations}
%
which are, for the whole system, constrained by the microcanonical restrictions
%
\begin{subequations}\label{eq:qGlobRest}
\begin{equation}\label{eq:qGlobRestN}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\right]\delta  V_M
    =\sum_{M=1}^{K}\mathcal{N}_{M}(t)\delta V_M=N,
\end{equation}
%
and
%
\begin{equation}\label{eq:qGlobRestE}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\epsilon_{n}\right]\delta V_M
    =\sum_{M=1}^{K}\mathcal{E}_M(t)\delta V_M=E. 
\end{equation}
\end{subequations}
%

Applying the variational method to $\mathcal{H}$, and using the
Lagrange multipliers
$\{\alpha_M\}$ and $\{\beta_M\}$, it is straightforward to obtain
(see also the discussion related to Eq.~(\ref{eq:deltaHpdeltafpj})):
%
\begin{equation}\label{eq:relation}
\ln \left(\frac{1\mp f_{Mn}(t)}{f_{Mn}(t)} \right)=-\alpha_M(t)-\beta_M(t) \epsilon_{n},
\end{equation}
%
and solving for $f_{Mn}(t)$ yields
%
\begin{equation}\label{eq:qfMn}
f_{Mn}(t)=\frac{1}{\exp\big(-\alpha_M(t)-\beta_M(t) \epsilon_{n}\big)\pm 1}.
\end{equation}
%
Thus, in this zero-order approximation, the form of equilibrium distribution functions are conserved.

%============================================================
\subsubsection{Properties of $\mathcal{H}$ for systems in equilibrium}
%============================================================
If the system is in equilibrium (we will show that it will also be valid at $t\to\infty$), the temperature becomes homogeneous throughout
the complete system. Also, the local number of particles, the local
energy, and the Lagrange multipliers do not depend on the cell number and they should be homogeneous,
this is 
%
\begin{subequations}\label{eq:qEqRestrictions}
\begin{equation}
   {\mathcal{N}}_M(t\to\infty)\equiv \bar{\mathcal{N}}=N/K,
\end{equation}
%
%
\begin{equation}
	{\mathcal{E}}_M(t\to\infty)\equiv \bar{\mathcal{E}}=E/K,
\end{equation}
\end{subequations}
%
%
\begin{subequations}\label{eq:qEqAlphaBeta}
\begin{equation}
	\alpha_M=\bar\alpha,\quad\forall M,
\end{equation}
%
and
\begin{equation}
	\beta_M=\bar\beta, \quad\forall M.
\end{equation}
\end{subequations}
%
Substituting Eqs.~(\ref{eq:qEqAlphaBeta}) into Eq.~(\ref{eq:qfMn}) yields the distribution
function of each cell in equilibrium to be
%
\begin{equation}\label{eq:qfneq}
    \bar f_{Mn}=\bar f_n =\frac{1}{\exp\big(-\bar\alpha-\bar\beta \epsilon_n\big)\pm 1},\quad\forall M.
\end{equation}
%
By means of the above equation, we can recover the distribution function and the entropy of a diluted quantum gas in equilibrium as follows. Setting $\bar\alpha=\mu/kT$ and $\bar\beta=-1/kT$,
and substituting them in Eq.~(\ref{eq:qfneq}), it renders the Fermi-Dirac and Bose-Einstein
distribution functions:
%
\begin{equation}\label{eq:qfneqtwo}
    \bar{f}_{n}=\frac{1}{\exp\big(\frac{{\epsilon_n}-\mu}{kT}\big)\pm 1},
\end{equation}
%
and substituting Eq.~(\ref{eq:qfneq}) into the negative of Eq.(\ref{eq:qHdef}), 
the entropy of a quantum ideal gas is obtained
%
\begin{eqnarray}\label{eq:varEntropyDef}
      S&=&\sum_{M=1}^{K}
        \sum_n  \left[
        		\left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
           \ln \left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
         \right]\nonumber \\
      &&\quad\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}
         \left(1 \mp \frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
         \right] \delta V_M.
\end{eqnarray}
%
This quantity is what we will refer to as ``variational entropy'',
and this name reflects the fact that it was obtained \textit{via}
the variational method.

@@

For quantum systems, we also accept the validity of the local equilibrium hypothesis
for every cell in the system. This allows to define non-homogeneous systems,
wherein thermodynamic quantities are well-defined on a cell-per-cell basis,
and in terms of the equilibrium properties, we have

@@

In the present context, $|\Delta_M|$ and $|\delta_M|$ are sufficiently large so they are not
fluctuations of the system, and sufficiently small so that the local equilibrium hypothesis is
valid for $t>0$ (we set $t_0=0$, and $t_0$ is the initial time at which the system is prepared).
Therefore, it is reasonable to re-write the distribution functions as:
%
\begin{equation}\label{eq:qFirstOrd}
   f_{Mn}(t)=\bar{f}_n(1+g_{Mn}(t)),\quad
   1\gg|g_{Mn}(t)|,
\end{equation}
%
from which it follows, by substituting Eq.~(\ref{eq:qFirstOrd}) into
Eq.~(\ref{eq:restrictionoutside}), that $\Delta_M$ and $\delta_M$ satisfy:
%
\begin{subequations}
\begin{equation}
    \Delta_M(t)=\sum_n \bar{f}_n g_{nM}
\end{equation}
%
and
%
\begin{equation}
	\delta_M(t)=\sum_n  \bar{f}_n g_{nM}\epsilon_n.
\end{equation}
\end{subequations}
%

@@

%\begin{proof}
To proof the quantum $H$-theorem, start by taking the time-derivative of Eq. (\ref{eq:qHdef}):
%
\begin{equation}\label{eq:deltaH}
   \frac{d \mathcal{H} (t)}{dt}= \sum_n \sum_{M=1}^{K} \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M.
\end{equation}
%
Subsequently, substitute Eq.~(\ref{eq:qFirstOrd}) in the above equation to obtain:
%
\begin{eqnarray}\label{eq:cambioH1}
    \frac{d\mathcal{H} (t)}{dt}&=&
      \sum_n \sum_{M=1}^{K} \bar{f}_{n}\ln \left[
        \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})}
      \right]\dot{g}_{nM} \delta V_M \nonumber \\
    &=&\sum_n \sum_{M=1}^{K} \bar{f}_n \left \{
      \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}
      -\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}
    \right \}\delta V_M.
\end{eqnarray}
%
Approximate the logarithmic terms, corresponding to
Fermi and Bose gases, through a Taylor series around $\bar f_n g_{nM}=0$ as:
%
\begin{subequations}\label{eq:qlnApprox}
\begin{equation}\label{eq:qlnApproxFermions}
	\ln[1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]
    	\approx \ln[1\mp\bar{f}_n]\mp\frac{\bar{f}_n}{1\mp\bar{f}_{n}} g_{nM}
\end{equation}
%
and
%
\begin{equation}\label{eq:qlnApproxBosons}
    \ln [\bar{f}_n+\bar{f}_n g_{nM}] \approx \ln [\bar{f}_n]+ g_{nM},
\end{equation}
\end{subequations}
%
respectively. Eq.~(\ref{eq:qlnApproxFermions}) is valid because, for non-extremely degenerated Fermi gases, 
$1-\bar{f}_n \gg \bar{f}_n|g_{nM}|$, and Eq.~(\ref{eq:qlnApproxBosons}) is fulfilled because,
for Boson gases,
$1+\bar{f}_n \gg \bar{f}_n |g_{nM}|$ when $\bar{f}_n \gg \bar{f}_n |g_{nM}|$.
